Let's implement a callback that prints 'Iteration `n`' (where `n` is the iteration number) for every batch in the training set and 'Done' when an epoch is finished. For that task we use the `luz_callback` function:

```{r, eval = FALSE}
print_callback <- luz_callback(
  name = "print_callback",
  initialize = function(message) {
    self$message <- message
  },
  on_train_batch_end = function() {
    cat("Iteration ", ctx$iter, "\n")
  },
  on_epoch_end = function() {
    cat(self$message, "\n")
  }
)
```

`luz_callback()` takes a named list of function as argument where the name indicate the moment at which the callback should be called. For instance `on_train_batch_end()` is called for every batch at the end of the training procedure and `on_epoch_end()` is called at the end of every epoch.

The returned value of `luz_callback()` is a function that initializes an instance
of the callback. Callbacks can have initialization parameters, like the name of a file
where you want to log the results. In that case, you can pass an `initialize` method when
creating the callback definition, and save these parameters to the `self` object. In the above example, the callback has a `message` parameter that is printed at the end of each epoch.

Once a callback is defined it can be passed to the `fit` function via the `callbacks` parameter, eg:

```{r, eval = FALSE}
fitted <- net %>%
  setup(...) %>%
  fit(..., callbacks = list(
    print_callback(message = "Done!")
  ))
```

Callbacks can be called in many different positions of the training loop, including combinations of them. Here's an overview of possible callback *breakpoints*:

    Start Fit
       - on_fit_begin
      Start Epoch Loop
         - on_epoch_begin
        Start Train
           - on_train_begin
          Start Batch Loop
             - on_train_batch_begin
              Start Default Training Step
                - on_train_batch_after_pred
                - on_train_batch_after_loss
                - on_train_batch_before_backward
                - on_train_batch_before_step
                - on_train_batch_after_step
              End Default Training Step:
             - on_train_batch_end
          End Batch Loop
           - on_train_end
        End Train
        Start Valid
           - on_valid_begin
          Start Batch Loop
             - on_valid_batch_begin
              Start Default Validation Step
                - on_valid_batch_after_pred
                - on_valid_batch_after_loss
              End Default Validation Step
             - on_valid_batch_end
          End Batch Loop
           - on_valid_end
        End Valid
          - on_epoch_end
      End Epoch Loop
       - on_fit_end
    End Fit

Every step market with `on_*` is a point in the training procedure that is available for callbacks to be called.

The other important part of callbacks is the `ctx` (context) object. See `help("ctx")` for details.

By default, callbacks are called in the same order as they were passed to `fit` (`predict` or `evaluate`), but you can provide a `weight` attribute that will control the order that it will be called. For example, if one callback has `weight = 10` and another has `weight = 1`, then the first one is called after the second one.
Callbacks that don't specify a `weight` attribute are considered `weight = 0`. A few built-in callbacks in luz already provide a weight value, for example the `?luz_callback_early_stopping`, since in general we want to run it as the last thing in the loop.
